Learning complex stochastic models with invertible neural networks: a likelihood-free Bayesian approach

### 2019-07-01 latex/paper_main.tex
Add seqRNA files

\title{Learning complex models with invertible neural networks: a likelihood-free Bayesian approach}

@@

\begin{figure}[H]
\centering
\begin{subfigure}{.49\textwidth}
    \includegraphics[width=\textwidth]{SIR_plot.png}
    \caption{Example SIR timeseries with $\beta=0.5$ and $\gamma=0.2$}
    \label{fig:Fig.6a}
\end{subfigure}
\begin{subfigure}{.49\textwidth}
    \includegraphics[width=\textwidth]{SIR_states.png}
    \caption{State-transition diagram of the SIR model}
    \label{fig:Fig.6b}
\end{subfigure}
\begin{subfigure}{.49\textwidth}
    \includegraphics[width=\textwidth]{SIR_scatter.png}
    \caption{Parameter recovery ($T=500$)}
    \label{fig:Fig.6c}
\end{subfigure}
\begin{subfigure}{.49\textwidth}
    \centering
    \begin{subfigure}{.49\textwidth}
    	\includegraphics[width=\textwidth]{SIR_metrics.png}
    	\caption{Performance over all $T$s}
    	\label{fig:Fig.6d}
	\end{subfigure}
	\begin{subfigure}{.49\textwidth}
    	\includegraphics[width=\textwidth]{SIR_variance.png}
    	\caption{Posterior variance over all $Ts$}
    	\label{fig:Fig.6e}
	\end{subfigure}
	\begin{subfigure}{.49\textwidth}
    	\includegraphics[width=\textwidth]{SIR_density.png}
    	\caption{Example posterior samples ($T=500$}
    	\label{fig:Fig.6f}
	\end{subfigure}
	\begin{subfigure}{.49\textwidth}
    	\includegraphics[width=\textwidth]{SIR_sbc.png}
    	\caption{Simulation-based calibration (SBC)}
    	\label{fig:Fig.6f}
	\end{subfigure}
\end{subfigure}
\caption[short]{Results on the stochastic SIR model. \textbf{(a)} Example SIR timeseries generated with $\beta=0.5$ and $\gamma=0.2$; \textbf{(b)} State transition diagram of the SIR model; \textbf{(c)} Parameter recovery depicting also NRMSE and $R^{2}$ metrics; \textbf{(d)} NRMSE and $R^{2}$ performance over all $T$s seen by the networks during training. We observe similar patterns as in the previous examples; \textbf{(e)} Posterior variances of the SIR parameters over all $T$s; \textbf{(e)} Example full posteriors of the parameters recovered from three test datasets; \textbf{(f)} Plots of the rank statistic, according to which the posterior of $\beta$ slightly overestimates the true mean.} \label{fig:Fig.6}
\end{figure}

@@

\subsection*{Example 4 - Single-Cell RNA Sequencing}
Single-cell RNA sequencing (scRNA-seq) is a method to uncover the dynamics of gene expression within single cells \cite{zappia2017splatter, ozsolak2011rna}. Simulation models for scRNA-seq attempt to mimic the process of sequencing real data samples by combining statistical and algorithmic procedures. The recently developed \textit{Splat} simulation model \cite{zappia2017splatter} implements a hierarchical model where the mean expression levels for multiple genes are samples from a Gamma distribution, and the number of times a gene has been sequenced in a cell is then sampled from a Poisson distribution. The output of the \textit{Splat} simulation is thus a matrix $\boldsymbol{X}$ of \textit{Gene x Cell} counts. Figure XXX depicts the \textit{Splat} algorithm for simulating scRNA-seq data.

As a last example for our method, we use simulations from the \textit{Splat} model and attempt to recover the data-generating parameters. We simulate $500000$  \textit{Gene x Cell} count matrices with different parameter settings (see \textbf{SI} for details of the simulation) and discard implausible simulations (e.g., matrices where more than $90\%$ of the counts are $0$). We train the networks for 100 epochs through the entire dataset and evaluate the performance on a separate validation set or matrices.

Figure XXX depicts the results...

>>>>

\title{Learning complex stochastic models with invertible neural networks: a likelihood-free Bayesian approach}

@@

\begin{figure}[H]
\centering
\begin{subfigure}{.49\textwidth}
    \includegraphics[width=\textwidth]{SIR_plot.png}
    \caption{Example SIR timeseries with $\beta=0.5$ and $\gamma=0.2$}
    \label{fig:Fig.6a}
\end{subfigure}
\begin{subfigure}{.49\textwidth}
    \includegraphics[width=\textwidth]{SIR_states.png}
    \caption{State-transition diagram of the SIR model}
    \label{fig:Fig.6b}
\end{subfigure}
\begin{subfigure}{.49\textwidth}
    \includegraphics[width=\textwidth]{SIR_scatter.png}
    \caption{Parameter recovery ($T=500$)}
    \label{fig:Fig.6c}
\end{subfigure}
\begin{subfigure}{.49\textwidth}
    \centering
    \begin{subfigure}{.49\textwidth}
    	\includegraphics[width=\textwidth]{SIR_metrics.png}
    	\caption{Performance over all $T$s}
    	\label{fig:Fig.6d}
	\end{subfigure}
	\begin{subfigure}{.49\textwidth}
    	\includegraphics[width=\textwidth]{SIR_variance.png}
    	\caption{Posterior variance over all $Ts$}
    	\label{fig:Fig.6e}
	\end{subfigure}
	\begin{subfigure}{.49\textwidth}
    	\includegraphics[width=\textwidth]{SIR_density.png}
    	\caption{Example posterior samples ($T=500$}
    	\label{fig:Fig.6f}
	\end{subfigure}
	\begin{subfigure}{.49\textwidth}
    	\includegraphics[width=\textwidth]{SIR_sbc.png}
    	\caption{Simulation-based calibration (SBC)}
    	\label{fig:Fig.6g}
	\end{subfigure}
\end{subfigure}
\caption[short]{Results on the stochastic SIR model. \textbf{(a)} Example SIR timeseries generated with $\beta=0.5$ and $\gamma=0.2$; \textbf{(b)} State transition diagram of the SIR model; \textbf{(c)} Parameter recovery depicting also NRMSE and $R^{2}$ metrics; \textbf{(d)} NRMSE and $R^{2}$ performance over all $T$s seen by the networks during training. We observe similar patterns as in the previous examples; \textbf{(e)} Posterior variances of the SIR parameters over all $T$s; \textbf{(f)} Example full posteriors of the parameters recovered from three test datasets; \textbf{(g)} Plots of the rank statistic, according to which the posterior of $\beta$ slightly overestimates the true mean.} \label{fig:Fig.6}
\end{figure}

@@

\subsection*{Example 4 - Single-Cell RNA Sequencing}
Single-cell RNA sequencing (scRNA-seq) is a method to uncover the dynamics of gene expression within single cells \cite{zappia2017splatter, ozsolak2011rna}. Simulation models for scRNA-seq attempt to mimic the process of sequencing real data samples by combining statistical and algorithmic procedures. The recently developed \textit{Splat} simulation model \cite{zappia2017splatter} implements a hierarchical model where the mean expression levels for multiple genes are samples from a Gamma distribution, and the number of times a gene has been sequenced in a cell is then sampled from a Poisson distribution. The output of the \textit{Splat} simulation is thus a matrix $\boldsymbol{X}$ of \textit{Gene x Cell} counts. \autoref{fig:Fig.7a} depicts the \textit{Splat} algorithm for simulating scRNA-seq data \cite{zappia2017splatter}.
\begin{table} 
\centering
\caption{Splat target parameters estimated by our method}
\begin{tabular}{lll}
Parameter & Name & Description\\
\midrule
$\alpha$ & Mean shape & Shape of the mean gene expression gamma distribution \\
$\beta$ & Mean rate & Rate of the mean gene expression gamma distribution \\
$\mu^{L}$ & Library size location & Location of the library size log-normal distribution \\
$\sigma^{L}$ & Library size scale & Scale of the library size log-normal distribution \\
$\pi^{O}$ & Outlier probability & Probability that a gene is an expression outlier \\
$\mu^{O}$ & Outlier location & Location of the expression outlier factor log-normal distribution \\
$\sigma^{O}$ & Outlier scale & Scale of the expression outlier factor log-normal distribution \\
$\phi$ & Common BCV & Common BCV dispersion across all genes \\
\bottomrule
\end{tabular}
\label{table:Table 1}
\end{table}
As a last example for our method, we use simulations from the \textit{Splat} model and attempt to recover the data-generating parameters. We simulate $500000$  \textit{Gene x Cell} count matrices with different parameter settings (see \textbf{SI} for details of the simulation) and discard implausible simulations (e.g., matrices where more than $50\%$ of the counts are $0$). The parameters considered for estimation, along with a short description, are listed in \autoref{table:Table 1}. The parameter estimation task is thus to recover $\boldsymbol{\theta} = (\alpha,\beta,\mu^{L},\sigma^{L},\pi^{O},\mu^{O},\sigma^{O},\phi)$ from an observed count matrix $\boldsymbol{X}$. We train the networks for 50 epochs through the entire dataset and evaluate the performance on a separate validation set of $500$ count matrices.
\begin{figure}[H]
\centering
\begin{subfigure}{.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{Splatter_plot.png}
    \caption{The \textit{Splat} simulation diagram. Adapted after \cite{zappia2017splatter}.}
    \label{fig:Fig.7a}
\end{subfigure}
\begin{subfigure}{.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{Splatter_scatter.png}
    \caption{Parameter recovery }
    \label{fig:Fig.7b}
\end{subfigure}
\caption[short]{Results on the \textit{Splat} scRNA simulation model. \textbf{(a)} The \textit{Splat} simulation diagram; \textbf{(b)} Parameter recovery.} \label{fig:Fig.7}
\end{figure}
The results on the scRNA example are depicted in \autoref{fig:Fig.7}. We observe excellent parameter recovery on the validation set. \textbf{TODO: Train the full method. These results are preliminary}.
